Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2024-07-06 manuscript/manuscript.tex
Working on discussion

\textbf{Conclusion:} x.
}

@@

This section discusses and summarizes the results from Section \ref{results}. Models for COD assignment are compared to physicians in Section \ref{discuss-compare}. Advantages and disadvantages of using GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models for assigning CODs are discussed in Sections \ref{discuss-adv} and \ref{discuss-disadv}. Limitations of the study are mentioned in Section \ref{discuss-limits}, while opportunities and future work are detailed in Sections \ref{discuss-opp} and \ref{discuss-futurework}.

\subsection{Comparison to Physician Assignment}\label{discuss-compare}

@@

\subsubsection{Natural Language Data}\label{discuss-adv-struct}

@@

\subsubsection{Computer Assisted Verbal Autopsy}\label{discuss-adv-assist}

The models in thi study can assist physicians in assigning CODs in a variety of ways. Similar to differential diagnoses, GPT and InSilicoVA models can offer more alternative COD assignments for physicians to consider,  \cite{barnettComparativeAccuracyDiagnosis2019}, which can potentially help lower the number of records with ill-defined causes or reduce disagreement in physicians. At the time of this study, running GPT-3.5 cost $\sim$\$1.6 USD (\$0.5 per one million tokens), GPT-4 cost $\sim$\$115 USD (\$30 per one million tokens), and InSilicoVA had no costs on 6939 physician agreed records \cite{openaiPricing2024}. These costs are likely lower than physicians (e.g. less than \$3 USD per house in India \cite{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005}). When physicians are unavailable, GPT and InSilicoVA models can be a cost-efficient alternative to code large amounts of records for population estimates of CODs. However, care needs to be taken to apply these models only for certain CODs where models perform well, such as in Table \ref{tab-perf-geq80}. In addition, these models can also help divert physician resources to cases that are more difficult to code or require more attention For example, physicians can simply check cases, such as road traffic injury where models performed well at 0.84 PCCC, while spending more time on cases that related to acute respiratory infections, where the models performed poorly at 0.62 PCCC (Figure \ref{fig-perf-cod-adult}).

@@

\subsubsection{Broad Cause of Death Assignment}\label{discuss-disadv-broad}

All models were evaluated for their performance in broad CGHR-10 COD categories as opposed to more specific ICD-10 codes. However, in practical cases, physicians assign more specific ICD-10 codes rather than broader COD categories. InterVA-5 and InSilicoVA assigned broader WHO VA codes, and were unable to assign ICD-10 codes, as the number of cases for specific ICD-10 codes may not have enough cases for training statistical models. GPT models are able to assign ICD-10 codes, but may possibly result in very low performance as even physicians do not agree completely on ICD-10 codes, and broader categories (CMEA-10 codes in Additional file 2) were used to assign equivalency or agreement.

\subsubsection{Reproducibility and Data Privacy}\label{discuss-disadv-reprod}

The GPT models in this study had the temperature parameter set to 0 for more reproducible results. A short experiment in Appendix \ref{app-reprod} revealed that GPT-3.5 assigns the same COD for the same record only more than 60\% of the time, based on repeated runs on a sample of 100 records. Although more extensive testing is needed, this suggests that GPT models do not always assign the COD for the same case on multiple runs, which may pose issues in reproducibility and reliability. For example, GPT models may achieve correct COD assignments solely due to random chance, but are difficult to test with large numbers (e.g. 10,000) of reruns due to costs (e.g. costs increased 10 fold per record when rerun 10 times). In addition, GPT models use prompts, which contain the narrative data, to assign CODs, which may be unknowingly collected and misused by companies (e.g. companies) or their users (e.g. malicious prompts) \cite{khowajaChatGPTNeedsSPADE2024,wuUnveilingSecurityPrivacy2024}. This poses issues with data privacy as sensitive data can be collected by these companies. In comparison, InterVA-5 and InSilicoVA are completely open source, and assign deterministic CODs with probabilities for each alternative COD, which offers more reproducible and reliable COD assignments despite lower performance overall. InterVA-5 and InSilicoVA can also be run without large computing infrastructure, which allows data to stay with the owner on their local computers.

@@

\subsubsection{Rare Causes, Disagreements, and Misclassifications}\label{discuss-limits-data}

Limited records for some causes of death. Did not explore adjudicated and reconciliated records.

@@

Did not explore effects of model parameters outside the default (e.g. ChatGPT temperature, etc).

Did not test multiple icd code outputs.

Did not explore response consistency with reruns.

@@

This section discusses research opportunities to improve GPT models for assigning CODs. Section \ref{discuss-opp-icd10} suggests comparing GPT models with physicians using ICD-10 codes. Section \ref{discuss-opp-peng} discusses the potential to improve GPT models with prompt engineering. Finally, an exploration of misclassified cases with domain knowledge is discussed in Section \ref{discuss-opp-misclass}.

\subsubsection{Comparisons with ICD-10}\label{discuss-opp-icd10}

For the scope of this study, performance evaluation was focused on broader CGHR-10 codes. However, comparisons should also be made on ICD-10 codes with equivalency using CMEA-10 codes to provide insights closer to what physicians assign in practice. Based on the results of this study, GPT, InterVA-5, and InSilicoVA models can assign accurate codes for some CODs (Table \ref{tab-perf-geq80}) using broader CGHR-10 codes, but may still rely on physicians to narrow these codes down to more specific ICD-10 codes.

\subsubsection{Prompt Engineering}\label{discuss-opp-peng}

Prompt engineering: adjust and optimize prompts, add in correctional instructions, adjudication, reconciliation, ICD-10/physician coding guide standards, and system prompt.

Sensitivity analysis: Test consistency of GPT responses with multiple reruns and parameter settings.

\subsubsection{Explore Misclassifications}\label{discuss-opp-misclass}

Look into incorrect assignments of CODs.

\subsection{Future Work}\label{discuss-futurework}

>>>>

\textbf{Conclusion:} GPT and InSilicoVA models were comparable to physicians for some CODs, but require further evaluation on reliability and larger datasets.
}

@@

This section discusses and summarizes the results from Section \ref{results}. Advantages and disadvantages of using GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models for assigning CODs are discussed in Sections \ref{discuss-adv} and \ref{discuss-disadv}. Limitations of the study are mentioned in Section \ref{discuss-limits}, while opportunities and future work are detailed in Section \ref{discuss-opp}.

This section identifies advantages of performant models for assigning CODs. Section \ref{discuss-adv-cod} details the strategic use of models for particular CODs. Section \ref{discuss-adv-efficient} details the resource efficiency of GPT and InSilicoVA models for assisting in physician COD assignment. Section \ref{discuss-adv-nlf} notes the strength of using natural language text in GPT models compared to structured data (e.g. questionnaires) for physician COD assignment.

\subsubsection{Comparable to Physicians for Some Causes of Death}\label{discuss-adv-cod}

@@

\subsubsection{Resource Efficient}\label{discuss-adv-efficient}

The models in this study can assist physicians in assigning CODs in a variety of ways due to low costs and speed of COD assignment. Similar to differential diagnoses, GPT and InSilicoVA models can offer more alternative COD assignments for physicians to consider, \cite{barnettComparativeAccuracyDiagnosis2019}, which can potentially help lower the number of records with ill-defined causes or reduce disagreement in physicians. At the time of this study, running GPT-3.5 cost $\sim$\$1.6 USD (\$0.5 per one million tokens), GPT-4 cost $\sim$\$115 USD (\$30 per one million tokens), and InSilicoVA had no costs on 6939 physician agreed records \cite{openaiPricing2024}. These costs are likely lower than physicians (e.g. less than \$3 USD per house in India \cite{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005}), while over 10,000 records can be coded in under a day. When physicians are unavailable, GPT and InSilicoVA models can be a cost-efficient alternative to code large amounts of records for population estimates of CODs. However, care needs to be taken to apply these models only for certain CODs where models perform well, such as in Table \ref{tab-perf-geq80}. In addition, these models can also help divert physician resources to cases that are more difficult to code or require more attention For example, physicians can simply check cases, such as road traffic injury where models performed well at 0.84 PCCC, while spending more time on cases that related to acute respiratory infections, where the models performed poorly at 0.62 PCCC (Figure \ref{fig-perf-cod-adult}).

\subsubsection{Natural Language Flexibility}\label{discuss-adv-nlf}

@@

@@

This section discusses the disadvantages of GPT models for COD assignment. Section \ref{discuss-disadv-reprod} identifies issues in reproducing GPT outputs for repeated runs on the same records and lack of up-to-date information, while Section \ref{discuss-disadv-repriv} discusses the resource intensive infrastructure required by GPT and its relation to data privacy.

\subsubsection{Reproducibility and Timeliness}\label{discuss-disadv-reprod}

The GPT models in this study had the temperature parameter set to 0 for more reproducible results. A short experiment in Appendix \ref{app-reprod} revealed that GPT-3.5 assigns the same COD for the same record only more than 60\% of the time, based on repeated runs on a sample of 100 records. Although more extensive testing is needed, this suggests that GPT models do not always assign the COD for the same case on multiple runs, which may pose issues in reproducibility and reliability. For example, GPT models may achieve correct COD assignments solely due to random chance, but are difficult to test with large numbers (e.g. 10,000) of reruns due to costs (e.g. costs increased 10 fold per record when rerun 10 times). In comparison, InterVA-5 and InSilicoVA are open source and free, and assign deterministic CODs with probabilities for each alternative COD, which offers more reproducible and reliable COD assignments despite lower performance overall. In addition, all models were trained on historical data up to particular points in time, which may not utilize the most up-to-date data available (e.g. latest online articles, social media, or books for GPT models).

\subsubsection{Infrastructure and Data Privacy}\label{discuss-disadv-repriv}

GPT-3.5 and GPT-4 models required large computing infrastructure to train and run, which was not possible to run on local computers, or setup due to costs. This poses issues with data privacy as sensitive data (e.g. identifying information) needs to be sent to company servers, which can be collected by companies (e.g. OpenAI) and misused \cite{taoOpeningPandoraBox2023}. For example, GPT models use prompts, which contain the narrative data, to assign CODs. The data in the promots may be unknowingly collected and misused by companies (e.g. companies) or their users (e.g. malicious prompts) \cite{khowajaChatGPTNeedsSPADE2024,wuUnveilingSecurityPrivacy2024}. In contrast, InterVA-5 and InSilicoVA can be run on local computers, which allows data to stay with the owner to protect data privacy, without reliance on external services.

@@

This section identifies limitations in this research in the context of GPT models. Section \ref{discuss-limits-icd10} identifies the omission of ICD-10 performance evaluations. Section \ref{discuss-limits-data} discusses data limitations, such as small samples, exclusion of disagreements, and exploration of incorrectly assigned cases. Section \ref{discuss-limits-model} mentions the need for parameter tuning and evaluation of consistency and multiple COD assignments.

\subsubsection{ICD-10 Performance Evaluation}\label{discuss-limits-icd10}

For the scope of this study, all models were evaluated for their performance in broad CGHR-10 COD categories as opposed to more specific ICD-10 codes. However, in practical cases, physicians assign more specific ICD-10 codes rather than broader COD categories. InterVA-5 and InSilicoVA assigned broader WHO VA codes, and were unable to assign ICD-10 codes, as the number of cases for specific ICD-10 codes may not have enough cases for training statistical models. GPT models were able to assign ICD-10 codes, but may possibly result in lower performance as even physicians do not agree completely on ICD-10 codes, and broader categories (CMEA-10 codes in Additional file 2) were used to assign equivalency or agreement.

\subsubsection{Small Samples, Disagreements, and Misclassifications}\label{discuss-limits-data}

Performance evaluations and analyses were omitted for CODs with less than 10 cases (e.g. congenital anomalies, suicide), in which models may perform well on, but have limited evidence to make conclusive insights on performance. In this study, only records where physicians agreed on the COD assignment were included in performance evaluations. However, the performance of GPT-4 only dropped slightly from 0.61 (agreed) to 0.53 (disagreed) PCCC while records almost doubled, which suggests that GPT models may potentially be able to assign agreeable CODs without reconciliation or adjudication for cases where physicians disagreed (Figure \ref{fig-perf-allvsagree}). In relation, GPT may be used to explore prompts that may possibly reconcile or adjudicate records in disagreement. Finally, an exploration of misclassifications was not conducted for this study, but may yield useful insights. For example, a non-comprehensive exploration was conducted in Appendix \ref{app-misclass} on misclassified GPT-4 records for neonatal infections, which found potential issues with the categorization of CGHR-10 codes, order of information in narratives, and guidelines of COD assignments. More records may also provide more conclusive evidence of model performance results as 6939 records may be inadequate evidence at the country level compared to more comprehensive studies (e.g. the million death study \cite{jhaProspectiveStudyOne2005}).

@@

GPT-3.5 and GPT-4 models used default parameters with the exception of setting the temperature to 0 for more consistent results. However, the temperature and other potential settings may be adjusted to possibly improve performance \cite{openaiOpenAIPlatformAPI2024}. In addition, GPT models may possibly produce inconsistent results even with the temperature set to 0 as discussed in Section \ref{discuss-disadv-reprod}. Thus, it is important to also test the reliability and consistency of GPT outputs to avoid coincidental results due to randomness \cite{johnsonAssessingAccuracyReliability2023,jangConsistencyAnalysisChatGPT2023,krishnaEvaluationReliabilityRepeatability2024}. InterVA-5 and InSilicoVA were able to provide multiple COD assignments with probabilities for each. GPT models can be prompted to produce more than one COD assignment, but was not explored in this study. This may be useful to evaluate the performance of suggested alternative COD assignments, which can help physicians reach agreement or reduce ill-defined assignments.

@@

This section discusses research opportunities to improve GPT models for assigning CODs. Section \ref{discuss-opp-peng} discusses the potential to improve GPT models with prompt engineering and exploration of misclassified records. Section \ref{discuss-opp-vaint} identifies an opportunity to integrate GPT, InterVA-5, and InSilicoVA models into VA systems for improving physician COD assignment.

\subsubsection{Prompt Engineering and Custom Models}\label{discuss-opp-peng}

Prompt engineering, the act of designing prompts to guide GPT models for better results \cite{wangPromptEngineeringHealthcare2024}, presents an important research opportunity that may improve performance of GPT models for COD assignment. As mentioned in Section \ref{discuss-limits-data} and exemplified in Appendix \ref{app-misclass}, an analysis of misclassified records may yield insights on adjusting prompts to assign more correct CODs in relation to physicians. In addition, subsequent prompts and examples can be used to add in correctional instructions and refine results, while additional information from the questionnaire and physician VA manuals may add better context \cite{meskoPromptEngineeringImportant2023}. Sensitivity analyses may be conducted to assess the effects on performance and consistency of results from modified prompts on a COD basis. GPT models may also be customized to specific domains or contexts, where objectives, behaviours, extra data, privacy, and evaluation tests can be adjusted to produce custom models that perform better in targeted domains or circumstances \cite{almasreDevelopmentEvaluationCustom2024}.

\subsubsection{Verbal Autopsy Integration}\label{discuss-opp-vaint}
